Circumnuclear Media and Accretion Rates of Quiescent Supermassive Black Holes

### 2015-04-06 ms.tex
fixed table caption
fixed 6.1

  conductive heat flux saturates to $F_{\rm cond}=5\phi \rho c_s^3$). 

@@

However, even if SN Ia are rare near the stagnation radius, SN Ia may cap the
SMBH accretion rate by periodically blowing gas out of the nucleus of
low mass galaxies.  Between successive SNe, stars release a
gaseous mass $M_{\rm g} \approx \eta M_{\star}\tau_{\rm Ia}/t_{\rm h}$
interior to the Ia radius, which is locally gravitationally bound to the SMBH
by an energy $E_{\rm bind} \sim M_{\rm g}\sigma_0^{2}\mathcal{F}^{2}$, where typically $\mathcal{F} 
\lesssim$ few.  From the above definitions we find that 
\be
\frac{E_{\rm Ia}}{E_{\rm bind}} \sim \frac{(v_{w}^{\rm Ia})^{2}}{2\mathcal{F}^{2}\sigma_0^{2}}.  
\ee
Hence, for low mass BHs with $\sigma_0 \ll \mathcal{F}v^{\rm Ia}_{w}/2$, SN Ia are capable of dynamically clearing out gas from radii $\sim r_{\rm Ia} \gtrsim r_{\rm s}$.  Even when heating is sufficiently weak that the stagnation radius formally exceeds $r_{\rm Ia}$, the SMBH accretion rate is still potentially limited to a value
\begin{eqnarray}
\frac{\dot{M}_{\rm Ia}}{\dot{M}_{\rm edd}} &\approx& \frac{\eta M_{\bullet}}{\dot{M}_{\rm edd} t_{\rm h}}\left(\frac{r_{\rm Ia}}{r_{\rm soi}}\right)^{2-\Gamma} \approx \nonumber \\
 && \begin{cases}
    4.5 \times 10^{-4} M_{\bullet,8}^{-1.33}(\tau_{\star}/t_{\rm h})^{-0.2}
   & \text{core} \\
    2.2 \times 10^{-4} \Mbheight^{-0.84}(\tau_{\star}/t_{\rm h})^{-0.6}   & \text{cusp},
  \end{cases}
  \label{eq:eddr_Ia}
\end{eqnarray}
obtained substituting the Ia radius $r_{\rm Ia}$ (eq.~[\ref{eq:rIa}]) for $r_{\rm s}$ in the derivation leading to our estimate of $\dot{M}$ (eq.~[$\ref{eq:eddr_analytic}$]).  Note, however, that for low mass SMBHs, $\dot{M}_{\rm Ia}$ generally exceeds the maximum accretion rate for a thermally stable flow $\dot{M}_{\rm TI}$ (eq.~[\ref{eq:Mdotmax}]), implying that dynamical `blow-out' from SNe Ia cannot alone prevent cooling instabilities in the nuclei of low mass galaxies.

@@

Figure \ref{fig:bh_growth} (top panel) shows with solid lines the
accretion rate as a function of black hole mass for the average star
formation heating, calculated from equation (\ref{eq:eddr_analytic})
using our results for the total wind heating
(Fig.~\ref{fig:mdot_mass}).  This average accretion rate increases
from $\dot{M} \sim 10^{-6}-10^{-5}M_{\odot}$ yr$^{-1}$ for low-mass
black holes ($M_{\bullet} \sim 10^{6}M_{\odot}$) to $\dot{M} \sim
10^{-4}M_{\odot}$ yr$^{-1}$ for $M_{\bullet} \sim 10^{9}M_{\sun}$.
These fall below the maximum thermally stable accretion rate,
$\dot{M}_{\rm TI}$ (yellow lines), except at the highest
black hole masses $\sim 10^{9}M_{\odot}$.

For low mass black holes, blow out from near the SN Ia radius caps the SMBH accretion rate at a value $\dot{M}_{\rm Ia} \sim 10^{-6}-10^{-5}M_{\odot}$ yr$^{-1}$ (green lines; eq.~[\ref{eq:eddr_Ia}]), which is however not low enough to prevent otherwise thermally-unstable flow at smaller radii (see discussion following eq.~[\ref{eq:eddr_Ia}]).  For higher mass black holes the stagnation resides close to the Ia radius, but the energy released by a single Ia is insufficient to unbind gas from the stellar bulge (although Ia heating contributes to the steady heating rate).

@@

Shown also for comparison in Figure \ref{fig:bh_growth} (top panel) is the accretion rate onto SgrA* calculated by \citet{Quataert:2004a}, as well as the range of $\dot{M}$ for the low-luminosity AGN NGC3115 derived through detailed modeling by \citealt{ShcherbakovWong+:2014a} (who find a range of accretion rates depending on the assumed model for thermal conductivity, although they do not include saturation of the conductive flux as considered in this work).  The SgrA* rate exceeds our estimate of $\dot{M}$ due to the average star formation history for the same $M_{\bullet}$ by almost two orders of magnitude.  This discrepancy can be understood by the fact that star formation in the central parsec of the Milky Way is not in steady-state: feedback is instead dominated by the stellar winds from the ring of young massive stars of mass $\sim 10^{4}M_{\odot}$ and estimated age $\sim 10$ Myr (e.g., \citealt{Schodel+07}).  Such a star formation history is better described by our `star burst' scenario, for which the value of $\eta \sim 100$ at $\tau_{\star} = 10^{7}$ yr (Fig.~\ref{fig:vwImp}) is a factor of $\sim 100$ times higher than the value of $\eta \sim 1$ predicted in the average SFH case (Fig.~\ref{fig:vwSources}, bottom panel).

>>>>

  conductive heat flux saturates to $F_{\rm cond}=5\phi \rho c_s^3$).  Entries for solutions found to be thermally unstable for all values of $\eta > 0.02$ are denoted as {\bf TI}.

@@

However, even if SN Ia are rare near the stagnation radius, SN Ia may cap the
SMBH accretion rate by periodically blowing gas out of the nucleus of
low mass galaxies.  Between successive SNe, stars release a
gaseous mass $M_{\rm g} \approx \eta M_{\star}\tau_{\rm Ia}/t_{\rm h}$
interior to the Ia radius, which is locally gravitationally bound to the SMBH
by an energy $E_{\rm bind} \sim M_{\rm g}\sigma_0^{2}\mathcal{F}^{2}$, where typically $\mathcal{F} 
\lesssim$ few.  From the above definitions we find that 
\be
\frac{E_{\rm Ia}}{E_{\rm bind}} \sim \frac{(v_{w}^{\rm Ia})^{2}}{2\mathcal{F}^{2}\sigma_0^{2}}.  
\label{eq:blowout}
\ee
Hence, for low mass BHs with $\sigma_0 \ll \mathcal{F}v^{\rm Ia}_{w}/2$, SN Ia are capable of dynamically clearing out gas from radii $\sim r_{\rm Ia} \gtrsim r_{\rm s}$.  Even when heating is sufficiently weak that the stagnation radius formally exceeds $r_{\rm Ia}$, the SMBH accretion rate is still potentially limited to a value
\begin{eqnarray}
\frac{\dot{M}_{\rm Ia}}{\dot{M}_{\rm edd}} &\approx& \frac{\eta M_{\bullet}}{\dot{M}_{\rm edd} t_{\rm h}}\left(\frac{r_{\rm Ia}}{r_{\rm soi}}\right)^{2-\Gamma} \approx \nonumber \\
 && \begin{cases}
    4.5 \times 10^{-4} M_{\bullet,8}^{-1.33}(\tau_{\star}/t_{\rm h})^{-0.2}
   & \text{core} \\
    2.2 \times 10^{-4} \Mbheight^{-0.84}(\tau_{\star}/t_{\rm h})^{-0.6}   & \text{cusp},
  \end{cases}
  \label{eq:eddr_Ia}
\end{eqnarray}
obtained substituting the Ia radius $r_{\rm Ia}$ (eq.~[\ref{eq:rIa}]) for $r_{\rm s}$ in the derivation leading to our estimate of $\dot{M}$ (eq.~[$\ref{eq:eddr_analytic}$]).  Note, however, that for low mass SMBHs, $\dot{M}_{\rm Ia}$ generally exceeds the maximum accretion rate for a thermally stable flow $\dot{M}_{\rm TI}$ (eq.~[\ref{eq:Mdotmax}]), implying that dynamical `blow-out' from SNe Ia cannot alone prevent cooling instabilities in the nuclei of low mass galaxies.

@@

Figure \ref{fig:bh_growth} (top panel) shows with solid lines the
accretion rate as a function of black hole mass for the average star
formation heating, calculated from equation (\ref{eq:eddr_analytic})
using our results for the total wind heating
(Fig.~\ref{fig:mdot_mass}).  This average accretion rate increases
from $\dot{M} \sim 10^{-6}-10^{-5}M_{\odot}$ yr$^{-1}$ for low-mass
black holes ($M_{\bullet} \sim 10^{6}M_{\odot}$) to $\dot{M} \sim
10^{-4}M_{\odot}$ yr$^{-1}$ for $M_{\bullet} \sim 10^{9}M_{\sun}$.
These fall below the maximum thermally stable accretion rate,
$\dot{M}_{\rm TI}$ (yellow lines), although they become closer at the highest
black hole masses $\sim 10^{9}M_{\odot}$.

For low mass black holes, blow out from SN Ia radius caps the SMBH accretion rate at a value $\dot{M}_{\rm Ia} \sim 10^{-6}-10^{-5}M_{\odot}$ yr$^{-1}$ (green lines; eq.~[\ref{eq:eddr_Ia}]), which is however not low enough to prevent otherwise thermally-unstable flow at smaller radii (see discussion following eq.~[\ref{eq:eddr_Ia}]).  On the other hand, for higher $M_{\bullet}$ the stagnation resides close to the Ia radius so Ia heating contributes to the steady gas heating rate, however the energy released by a single Ia is insufficient to unbind gas from the stellar bulge (eq.~[\ref{eq:blowout}]).

@@

Shown also for comparison in Figure \ref{fig:bh_growth} (top panel) is the accretion rate onto SgrA* calculated by \citet{Quataert:2004a}.  Also show is the range of $\dot{M}$ for the low-luminosity AGN NGC3115 derived through detailed modeling by \citealt{ShcherbakovWong+:2014a}, who find a range of accretion rates depending on the assumed model for thermal conductivity (although we expect this range would be smaller had they included saturation of the conductive flux).  The SgrA* rate exceeds our estimate of $\dot{M}$ due to the average star formation history for the same $M_{\bullet}$ by almost two orders of magnitude.  This discrepancy can be understood by the fact that star formation in the central parsec of the Milky Way cannot be described as steady-state: feedback is instead dominated by the stellar winds from the ring of young massive stars of mass $\sim 10^{4}M_{\odot}$ and estimated age $\sim 10$ Myr (e.g., \citealt{Schodel+07}).  Such a star formation history is better described by our `star burst' scenario, for which the value of $\eta \sim 100$ at $\tau_{\star} = 10^{7}$ yr (Fig.~\ref{fig:vwImp}) is a factor of $\sim 100$ times higher than the value of $\eta \sim 1$ predicted in the average SFH case (Fig.~\ref{fig:vwSources}, bottom panel).
